---
title: "Group 125"
output: html_notebook
---

## Import all the libraries as and when needed

```{r}
library(psych) #use for summary
library(ggplot2) #use for plotting
library(GGally)
library(sf)
library(dplyr)
library(spdep)
library(lmtest)
library(texreg)
library(spatialreg)
library(viridis)
```

```{r}
load("data_125.RData")
```

## Question 1: Descriptive Statistics of the County data dictionary

```{r}
# Understand the structure of the county_leukemia_125 data dictionary
str(county_leukemia_125)

# Summarize each column to get descriptive statistics for the whole county
summary(county_leukemia_125[, c("id")]) # Since ids are categorical values it's not helpful to do statistics over such column.
# Hence we rather described if the id is unique or not so that they are no duplicates
length(unique(county_leukemia_125$id)) == nrow(county_leukemia_125) # Output is true, hence all the district ids are unique

summary(county_leukemia_125[, c("leuk_cases")])
#describe(county_leukemia_125[, c("leuk_cases")])

# Give the understanding the AverageAge is a categorical variable with 3 levels, we wanted to describe the data via these categories
summary(county_leukemia_125[, c("AverageAge")])

# Summarize GreenNeighnorhoodIndex
summary(county_leukemia_125[, c("GreenNeighborhoodIndex")])

#Summarize cars per household
summary(county_leukemia_125[, c("CarsPerHH")])

# Summarize the population growth
summary(county_leukemia_125[, c("PctPopGrowth")])

# Summarize the fracing locations statistics
loc_num <- 1:10

for (num in loc_num) {
  location <- paste0("Dist2FracLoc_", num) 
  print(summary(county_leukemia_125[, c(location)]))
}
```

Analysis: It shows that the data is has 500 samples with 20 variables. There are various datatypes like integer, numerical, Factor, etc.

## Question 2: Plot your variables using maps and pairs plots

### Maps of all variables

```{r}
#--------------------------Generating Maps---------------------------
print("Create basic plot for all the columns")
plot(county_leukemia_125,max.plot = 19)

#---------------------------------------------------------------------
print("Create a theme map for luekemia cases in unscaled manner")
ggplot(county_leukemia_125) +
  geom_sf(aes(fill = leuk_cases)) +
  geom_sf(data = fracking_locations_125,
          color = "red",
          size = 3) +
  scale_fill_gradient(low = "lightyellow", high = "orange", name = "Luekmia Cases (Unscaled)") + theme_minimal()
#---------------------------------------------------------------------
print("Create a theme map for Green Neighborhood Index")
ggplot(county_leukemia_125) +
  geom_sf(aes(fill = GreenNeighborhoodIndex)) +
  geom_sf(data = fracking_locations_125,
          color = "red",
          size = 3) +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen", name = "Green Neighborhood Index") + theme_minimal()
#---------------------------------------------------------------------
print("Create a theme map for Average Age")
ggplot(county_leukemia_125) +
  geom_sf(aes(fill = AverageAge)) +
  geom_sf(data = fracking_locations_125,
          color = "red",
          size = 3) + theme_minimal()
#---------------------------------------------------------------------
print("Create a theme map for Cars per household")
ggplot(county_leukemia_125) +
  geom_sf(aes(fill = CarsPerHH)) +
  geom_sf(data = fracking_locations_125,
          color = "black",
          size = 3) + scale_fill_gradient(low = "lightblue", high = "red", name = "Cars per household")+theme_minimal()
#---------------------------------------------------------------------
print("Create a theme map for population growth(%)")
ggplot(county_leukemia_125) +
  geom_sf(aes(fill = PctPopGrowth)) +
  geom_sf(data = fracking_locations_125,
          color = "red",
          size = 3)+ scale_fill_gradient(low = "yellow", high = "purple", name = "Population Growth (%)") + theme_minimal()

#---------------------------------------------------------------------
print("Create a theme map for population density")
ggplot(county_leukemia_125) +
  geom_sf(aes(fill = PopDens)) +
  geom_sf(data = fracking_locations_125,
          color = "red",
          size = 3) + scale_fill_gradient(name = "Population Density")+ theme_minimal()

```

### Map of scaled cases

```{r}

# Since the population density is relative so will be the cases we also scaled the cases with population density then used min-max normalization to normalize the values between 0 and 1 million.

# Checking if fracking locations need to be transformed for county geometry
fracking_locations_125_adjusted <- st_transform(fracking_locations_125, st_crs(county_leukemia_125))

# Scale the Luekemia cases

# Add column called cases_scaled to see how the cases are relatively distributed in the county
county_leukemia_125$cases_scaled <- county_leukemia_125$leuk_cases/county_leukemia_125$PopDens

# Use min-max normalization to scale the cases between 0 to 1_000_000
county_leukemia_125$cases_scaled <- 
(
  (county_leukemia_125$cases_scaled - 
  min(county_leukemia_125$cases_scaled)) / 
  (max(county_leukemia_125$cases_scaled) - 
  min(county_leukemia_125$cases_scaled))
) * 1000000

# Plot the scaled cases to see for patterns with respect to fracking sites. 
ggplot(county_leukemia_125) +
  geom_sf(aes(fill = cases_scaled), size = 1) +
  geom_sf(data = fracking_locations_125_adjusted,
          color = "red",
          size = 2)+
  scale_fill_gradient(low = "lightyellow", high = "violet", name = "Luekmia Cases (Scaled)")+
  theme_minimal()
```

### Pair plot of variables

#### Add median and mean of Fracking location sites

```{r}

# Calculate median for Dist2FracLoc_1 to Dist2FracLoc_10
county_leukemia_125 <- county_leukemia_125 %>%
  rowwise() %>%
  mutate(FrackingLocationsDistance_median = median(c_across(starts_with("Dist2FracLoc_"))))

# Calculate mean for Dist2FracLoc_1 to Dist2FracLoc_10
county_leukemia_125 <- county_leukemia_125 %>%
  rowwise() %>%
  mutate(FrackingLocationsDistance_mean = mean(c_across(starts_with("Dist2FracLoc_"))))
```

#### Create Pair Plot

```{r}
selected_Vars <- county_leukemia_125[, c("leuk_cases","cases_scaled","FrackingLocationsDistance_median","FrackingLocationsDistance_mean", "AverageAge", "GreenNeighborhoodIndex", "CarsPerHH", "PctPopGrowth")]

custom_labels <- c(
  leuk_cases = "Leukemia Cases",
  cases_scaled = "Scaled Cases",
  FrackingLocationsDistance_median = "Median Fracking Dist",
  FrackingLocationsDistance_mean = "Mean Fracking Dist",
  AverageAge = "Average Age",
  GreenNeighborhoodIndex = "Green Neighborhood Index",
  CarsPerHH = "Cars per Household",
  PctPopGrowth = "Population Growth"
)

selected_vars <- st_drop_geometry(selected_Vars)
p <- ggpairs(selected_vars, 
        lower = list(continuous = wrap("points", color = "purple")), 
        diag = list(continuous = wrap("densityDiag", fill = "lightgreen")), 
        upper = list(continuous = wrap("cor", color="black")),
        columnLabels = custom_labels,
        )

ggsave("ggpairs_plot.png", plot = p, width = 12, height = 10, dpi = 300)
```

## Question 3: Is the data generating process of leukemia cases random or can you find and numerically prove a spatial clustering?

```{r}
# Checking if the geomtry is still valid
class(county_leukemia_125)
st_is_valid(county_leukemia_125)
```

```{r}

neighbors <- poly2nb(county_leukemia_125)
weights <- nb2listw(neighbors)

# Calculate Moran's I
moran_test <- moran.test(county_leukemia_125$leuk_cases, weights)
print(moran_test)

# Calculate Moran's I
moran_test <- moran.test(county_leukemia_125$leuk_cases, weights)
print(moran_test)

# Calculate Moran's I
moran_test <- moran.test(county_leukemia_125$cases_scaled, weights)
print(moran_test)

#-------------------------------------------------------------------------------------------------------
# Visualize the clustering:
# Ensure variable and spatial weights are valid
luekemia_cases <- county_leukemia_125$leuk_cases
weights <- nb2listw(poly2nb(county_leukemia_125))

# Run Local Moran's I
local_moran <- localmoran(luekemia_cases, weights)

# Check the structure of the output
str(local_moran)

# Add results to your spatial dataset
county_leukemia_125$local_I <- local_moran[, "Ii"]      # Local Moran's I statistic
county_leukemia_125$z_score <- local_moran[, "Z.Ii"]    # Z-score
county_leukemia_125$p_value <- 2 * (1 - pnorm(abs(local_moran[, "Z.Ii"])))  # Compute p-values if missing

county_leukemia_125$cluster <- ifelse(county_leukemia_125$p_value < 0.05 & county_leukemia_125$local_I > 0, "Hotspot",
               ifelse(county_leukemia_125$p_value < 0.05 & county_leukemia_125$local_I < 0, "Coldspot", "Not Significant"))

# Plot hotspots and cold spots
ggplot(data = county_leukemia_125) +
  geom_sf(aes(fill = cluster)) +  # Color by cluster type
  scale_fill_manual(values = c("Hotspot" = "orange", "Coldspot" = "green", "Not Significant" = "grey")) +
  geom_sf(aes(fill = cluster), color = "black", size = 0.1)+
  geom_sf(data = fracking_locations_125_adjusted,
          color = "red",
          size = 2)+
  theme_minimal() +
  labs(title = "LISA Analysis: Hotspots and Cold Spots",
       fill = "Cluster Type")
```

## Question 4: Try to identify a significant correlation between the leukemia cases and explanatory variables using appropriate models.

### Create Linear models for variables with Leukemia cases as target variable

```{r}

# Create linear models for explanatory variables
fdm_lm <- lm(county_leukemia_125$leuk_cases ~ county_leukemia_125$FrackingLocationsDistance_median, county_leukemia_125)
avg_age_lm <- lm(county_leukemia_125$leuk_cases ~ county_leukemia_125$AverageAge, county_leukemia_125)
green_ng_lm <- lm(county_leukemia_125$leuk_cases ~ county_leukemia_125$GreenNeighborhoodIndex, county_leukemia_125)
pct_pop_growth_lm <- lm(county_leukemia_125$leuk_cases ~ county_leukemia_125$PctPopGrowth, county_leukemia_125)

# Combinations of explanatory variables
fdm_avg_age_lm <- lm(county_leukemia_125$leuk_cases ~ county_leukemia_125$AverageAge + county_leukemia_125$FrackingLocationsDistance_median, county_leukemia_125)
fdm_green_ng_lm <- lm(county_leukemia_125$leuk_cases ~ county_leukemia_125$GreenNeighborhoodIndex + county_leukemia_125$FrackingLocationsDistance_median, county_leukemia_125)
fdm_pct_pop_growth_lm <- lm(county_leukemia_125$leuk_cases ~ county_leukemia_125$PctPopGrowth + county_leukemia_125$FrackingLocationsDistance_median, county_leukemia_125)

# Combination of all explanatory variables
fdm_avg_age_green_ng_pct_pop_growth_lm <- lm(county_leukemia_125$leuk_cases ~ county_leukemia_125$GreenNeighborhoodIndex + county_leukemia_125$FrackingLocationsDistance_median +county_leukemia_125$AverageAge+county_leukemia_125$PctPopGrowth, county_leukemia_125)

screenreg(
  list(fdm_lm,
       avg_age_lm,
       green_ng_lm,
       pct_pop_growth_lm,
       fdm_avg_age_lm,
       fdm_green_ng_lm,
       fdm_pct_pop_growth_lm,
       fdm_avg_age_green_ng_pct_pop_growth_lm
    ),
  digits = 2
)
```

### Plot Residuals vs Fitted Values

```{r}

plot_data <- data.frame(
  FittedValues = fdm_lm$fitted.values,
  Residuals = resid(fdm_lm)
)

# Create the ggplot
ggplot(plot_data, aes(x = FittedValues, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(
    x = "Fitted Values",
    y = "Residuals",
    title = "Residuals vs Fitted Values"
  ) +
  theme_minimal()

plot_data <- data.frame(
  FittedValues = avg_age_lm$fitted.values,
  Residuals = resid(avg_age_lm)
)

# Create the ggplot
ggplot(plot_data, aes(x = FittedValues, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(
    x = "Fitted Values",
    y = "Residuals",
    title = "Residuals vs Fitted Values"
  ) +
  theme_minimal()

plot_data <- data.frame(
  FittedValues = green_ng_lm$fitted.values,
  Residuals = resid(green_ng_lm)
)

# Create the ggplot
ggplot(plot_data, aes(x = FittedValues, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(
    x = "Fitted Values",
    y = "Residuals",
    title = "Residuals vs Fitted Values"
  ) +
  theme_minimal()

plot_data <- data.frame(
  FittedValues = pct_pop_growth_lm$fitted.values,
  Residuals = resid(pct_pop_growth_lm)
)

# Create the ggplot
ggplot(plot_data, aes(x = FittedValues, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(
    x = "Fitted Values",
    y = "Residuals",
    title = "Residuals vs Fitted Values"
  ) +
  theme_minimal()
#Checking fitness for combination of values------------------------------------
plot_data <- data.frame(
  FittedValues = fdm_avg_age_lm$fitted.values,
  Residuals = resid(fdm_avg_age_lm)
)

# Create the ggplot
ggplot(plot_data, aes(x = FittedValues, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(
    x = "Fitted Values",
    y = "Residuals",
    title = "Residuals vs Fitted Values"
  ) +
  theme_minimal()

plot_data <- data.frame(
  FittedValues = fdm_green_ng_lm$fitted.values,
  Residuals = resid(fdm_green_ng_lm)
)

# Create the ggplot
ggplot(plot_data, aes(x = FittedValues, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(
    x = "Fitted Values",
    y = "Residuals",
    title = "Residuals vs Fitted Values"
  ) +
  theme_minimal()

plot_data <- data.frame(
  FittedValues = fdm_pct_pop_growth_lm$fitted.values,
  Residuals = resid(fdm_pct_pop_growth_lm)
)

# Create the ggplot
ggplot(plot_data, aes(x = FittedValues, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(
    x = "Fitted Values",
    y = "Residuals",
    title = "Residuals vs Fitted Values"
  ) +
  theme_minimal()

plot_data <- data.frame(
  FittedValues = fdm_avg_age_green_ng_pct_pop_growth_lm$fitted.values,
  Residuals = resid(fdm_avg_age_green_ng_pct_pop_growth_lm)
)

# Create the ggplot
ggplot(plot_data, aes(x = FittedValues, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(
    x = "Fitted Values",
    y = "Residuals",
    title = "Residuals vs Fitted Values"
  ) +
  theme_minimal()


```

### Check with BP test

```{r}
bptest(fdm_avg_age_green_ng_pct_pop_growth_lm)
```

## Question 5: Account for any spatial process that may underlie the coherence identified in 4.

### Do Moran Test for Leukemia Cases and the Residuals of a best model to check spatial autocorrelation.

```{r}

# Define neighbors and spatial weights
neighbors <- poly2nb(county_leukemia_125)          # Neighbors based on polygons
weights <- nb2listw(neighbors, style = "W")        # Spatial weights matrix

# Moran's I test for the dependent variable
moran_test <- moran.test(county_leukemia_125$leuk_cases, weights)
print(moran_test)

# Extract residuals from our best model
fdm_avg_age_green_ng_pct_pop_growth_lm_residuals <- residuals(fdm_avg_age_green_ng_pct_pop_growth_lm)

# Moran's I test for residuals
moran_test_residuals <- moran.test(fdm_avg_age_green_ng_pct_pop_growth_lm_residuals, weights)
print(moran_test_residuals)
```

### Significant Spatial Autocorrelation Found: Apply Spatial Regression Models

```{r}
# Both moran tests show high autocorrelation, we use spatial regression model now

# Fit a spatial lag model
slm_model <- lagsarlm(county_leukemia_125$leuk_cases ~ county_leukemia_125$FrackingLocationsDistance_median + county_leukemia_125$AverageAge + county_leukemia_125$GreenNeighborhoodIndex + county_leukemia_125$PctPopGrowth, data = county_leukemia_125, listw = weights)
summary(slm_model)

# Fit a spatial error model
sem_model <- errorsarlm(county_leukemia_125$leuk_cases ~ county_leukemia_125$FrackingLocationsDistance_median + county_leukemia_125$AverageAge + county_leukemia_125$GreenNeighborhoodIndex + county_leukemia_125$PctPopGrowth, data = county_leukemia_125, listw = weights)
summary(sem_model)

AIC(fdm_avg_age_green_ng_pct_pop_growth_lm, slm_model, sem_model)

moran.test(residuals(slm_model), weights)
moran.test(residuals(sem_model), weights)

# From the AIC and moran test we can conclude that the SEM model best fits and removes the spatial autocorrelation
```

## Question 6: Is there a relationship between the proximity to a fracking site and the number of leukemia cases? Underpin and justify your statement with data!

```{r}
ggplot(data = county_leukemia_125, aes(x = FrackingLocationsDistance_median, y = leuk_cases)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Relationship Between Fracking Proximity and Leukemia Cases",
       x = "Proximity to Fracking Sites (Median Distance)",
       y = "Number of Leukemia Cases")

```

```{r}
# End of Assignment
```
